------------------------------------------------------------------------

```{r setup, include=FALSE}
all_times <- list()  # store the time for each chunk
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res <- difftime(Sys.time(), now, units = "secs")
      all_times[[options$label]] <<- res
    }
  }
}))
knitr::opts_chunk$set(
  tidy = TRUE,
  tidy.opts = list(width.cutoff = 95),
  message = FALSE,
  warning = FALSE,
  time_it = TRUE
)
```

# Setup the Seurat Object

```{r init}
library(dplyr)
library(Seurat)
library(patchwork)

# Load the PBMC dataset
pbmc.data <- Read10X(data.dir = "fresh_68k_pbmc_donor_a_filtered_gene_bc_matrices/filtered_matrices_mex/hg19/")
# Initialize the Seurat object with the raw (non-normalized data).
pbmc <- CreateSeuratObject(counts = pbmc.data, project = "pbmc68k", min.cells = 3, min.features = 200)
pbmc

```

<details>


## QC and selecting cells for further analysis

```{r mito, fig.height=7, fig.width=13}
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
```

<details>

<summary>**QC metrics**</summary>

```{r qc, fig.height=7, fig.width=13}
# Show QC metrics for the first 5 cells
head(pbmc@meta.data, 5)
```

</details>
\

Visualize QC metrics, and use these to filter cells.

-   filter cells that have unique feature counts over 2,500 or less than 200
-   filter cells that have \>5% mitochondrial counts

```{r qc2, fig.height=7, fig.width=13}

#Visualize QC metrics as a violin plot
VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)

plot1 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt") 
plot2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") 
plot1 + plot2

pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
```

------------------------------------------------------------------------

# Normalizing the data


```{r normalize}
pbmc <- NormalizeData(pbmc, normalization.method = "LogNormalize", scale.factor = 1e4)
```

# Identification of highly variable features (feature selection)

```{r var_features, fig.height=5, fig.width=11}

pbmc <- FindVariableFeatures(pbmc, selection.method = 'mvp', nfeatures = 1000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(pbmc), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(pbmc)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2
```

# Scaling the data

```{r regressvarmt, fig.height=7, fig.width=11, results='hide',eval = FALSE}
all.genes <- rownames(pbmc)
pbmc <- ScaleData(pbmc, features = all.genes, vars.to.regress = "percent.mt")
```


# Perform linear dimensional reduction


```{r pca,results='hide'}
pbmc <- RunPCA(pbmc, features = VariableFeatures(object = pbmc))
```


```{r pca_viz, message=TRUE}
# Examine and visualize PCA results a few different ways
print(pbmc[['pca']], dims = 1:5, nfeatures = 5)
VizDimLoadings(pbmc, dims = 1:2, reduction = 'pca')
DimPlot(pbmc, reduction = 'pca')
```


```{r single-heatmap}
DimHeatmap(pbmc, dims = 1, cells = 500, balanced = TRUE)
```

```{r multi-heatmap, fig.height=15, fig.width=9}
DimHeatmap(pbmc, dims = 1:50, cells = 500, balanced = TRUE)
```

# Determine the 'dimensionality' of the dataset

```{r elbow_plot, fig.height=6, fig.width=10}
ElbowPlot(pbmc, ndims = 50)
```

```{r jackstraw, fig.height=6, fig.width=10}
pbmc <- JackStraw(pbmc, num.replicate = 100)
pbmc <- ScoreJackStraw(pbmc, dims = 1:20) 
```

The `JackStrawPlot()` function provides a visualization tool for comparing the distribution of p-values for each PC with a uniform distribution (dashed line). 'Significant' PCs will show a strong enrichment of features with low p-values (solid curve above the dashed line). In this case it appears that there is a sharp drop-off in significance after the first 10-12 PCs.

```{r jsplots, fig.height=6, fig.width=10}
JackStrawPlot(pbmc, dims = 1:20)
```


# Cluster the cells

```{r cluster, fig.height=5, fig.width=7}
pca_embeddings <- Embeddings(pbmc, reduction = "pca")[, 1:50]
set.seed(123)  
km_res <- kmeans(pca_embeddings, centers = 10)
pbmc$kmeans_cluster <- as.factor(km_res$cluster)
Idents(pbmc) <- pbmc$kmeans_cluster
```

# Run tSNE

```{r tsne, fig.height=5, fig.width=7}
pbmc <- RunTSNE(pbmc, dims = 1:50)
```

# Visualize tSNE

```{r tsne, fig.height=5, fig.width=7}
DimPlot(pbmc, reduction = "tsne", group.by = "kmeans_cluster", label = TRUE) + ggtitle("t-SNE: K-means Clusters (k=10)")
```

```{r saveobject, eval=FALSE}
saveRDS(pbmc, file = "pbmc_68k.rds")
```



## Subclustering within Each K-means Cluster


```{r}
library(cluster)
library(factoextra)
Idents(pbmc) <- pbmc$kmeans_cluster 
pbmc$subcluster <- NA

for (cluster_id in levels(pbmc$kmeans_cluster)) {
  cat("Processing cluster", cluster_id, "\n")
  
  subset_cells <- subset(pbmc, idents = cluster_id)
  
  if (ncol(subset_cells) < 10) {
    message("Too few cells in cluster ", cluster_id, ", skipping...")
    next
  }
  
  subset_pca <- Embeddings(subset_cells, reduction = "pca")[, 1:50]
  
  wss <- sapply(1:6, function(k) {
    kmeans(subset_pca, centers = k, nstart = 10)$tot.withinss
  })
  optimal_k <- which.min(diff(diff(wss))) + 1
  optimal_k <- max(2, min(optimal_k, 5))
  
  # Run K-means
  km_sub <- kmeans(subset_pca, centers = optimal_k, nstart = 10)
  
  pbmc$subcluster[Cells(subset_cells)] <- paste0(cluster_id, "_", km_sub$cluster)
}

```

```{r tsne-subclusters, fig.height=15, fig.width=9}
pbmc$kmeans_cluster <- as.factor(pbmc$kmeans_cluster)
pbmc$subcluster <- as.factor(pbmc$subcluster)
library(ggplot2)

# Extract t-SNE coordinates
tsne_coords <- as.data.frame(Embeddings(pbmc, "tsne"))
tsne_coords$subcluster <- pbmc$subcluster
tsne_coords$kmeans_cluster <- pbmc$kmeans_cluster
ggplot(tsne_coords, aes(x = tSNE_1, y = tSNE_2, color = subcluster)) +
  geom_point(size = 0.5, alpha = 0.8) +
  facet_wrap(~ kmeans_cluster, ncol = 5) +
  theme_minimal() +
  labs(title = "t-SNE: Subclusters per Original K-means Cluster",
       color = "Subcluster") +
  theme(strip.text = element_text(size = 10, face = "bold"),
        legend.position = "right",
        panel.spacing = unit(1, "lines"))

```
